HAMdetector: A Bayesian regression model that integrates information to detect HLA-associated mutations

### 2021-02-16 docs/main.tex
remove wrong positivity constraint from phylogeny gamma
add b_epi to full model specification
add genes to dataset list

The logit transform is used because is cancels out with the logistic inverse link function. This means that the phylogeny information acts as a baseline in absence of any HLA effects. Note that we also include an additional parameter \(\gamma\), that is constrained to be positive. This helps because the inferred probabilities \(P(y_{ik}=1|\Psi)\) might not necessarily reflect the true underlying phylogenetic signal, for example because the phylogenetic tree does not match the observed data well enough.

@@

\begin{equation}
  \label{eq:full}
  \begin{aligned}
   y_{ik} &\sim \text{Bernoulli}(\theta_{ik}) &\\
   \theta_{ik} &=
        \text{logit}^{-1}(\beta_{0_{k}} + \gamma_{k}\text{logit}(P(y_{ik}=1|\Psi)) & \\
        \hspace{-6cm} &~~~~ +\sum_{k=1}^{D}X_{ij}\beta_{\text{HLA}_{jk}}) & \\
    \beta_{0_{k}} &\sim \text{Normal}(0, 100^2) & \text{(*)}\\
   \gamma_{k} &\sim \text{Normal}(\mu_{\text{phy}}, \sigma_{\text{phy}}^2) & \\
    \mu_{\text{phy}} &\sim \text{Normal}(1, 1) & \text{(*)} \\
   \sigma_{\text{phy}} &\sim \text{Normal}^{+}(0, 0.5) & \text{(*)} \\
   \beta_{\text{HLA}_{jk}} &\sim \text{Normal}(0, \tau_{k}^{2}\tilde{\lambda}^{2}_{jk}) & \\
   \tilde{\lambda}^{2}_{jk} &= \frac{c_{k}^{2}\lambda_{jk}^{2}}{c_{k}^{2} + \tau_{k}^{2}\lambda_{jk}^{2}} & \\
    c_{k}^2 &\sim \text{Inv-Gamma}(3.5, 3.5) & \text{(*)} \\
   \lambda_{jk} &\sim \text{Cauchy}^{+}(0, \sigma_{j}^2\text{exp}(Z_{jk}\beta_{\text{epi}})) & \\
   \tau_{k} &\sim \text{Cauchy}^{+}(0, \tau_{0k}) & \text{(*)} \\
   \tau_{0k} &= \frac{10}{D-10}\frac{2}{\sqrt{N}} & \\
  \end{aligned}
\end{equation}

@@

\begin{itemize}
  \item A large HIV dataset consisting of 1383 sequences, that were kindly provided by Zabrina Brumme and collaborators and were also used in the Phylogenetic Dependency Network study~\citep{Carlson2008}.
  \item A set of 351 HIV sequences mostly spanning the Pol gene from the Arevir database~\citep{Roomp2006}.
  \item A set of 544 Hepatitis-B-Virus sequences that were kindly provided by Jörg Timm.
  \item A set of 104 Hepatitis-D-Virus sequences that were kindly provided by Michael Roggendorf. \hl{DaHo: bitte von Hadi geeignetes Zitat erfragen}
  \item A set of 41 HIV sequences that were kindly provided by Rongge Yang. 
\end{itemize}

\hl{DaHo: zu jedem Datensatz die Proteine nennen}

>>>>

The logit transform is used because is cancels out with the logistic inverse link function. This means that the phylogeny information acts as a baseline in absence of any HLA effects. Note that we also include an additional parameter \(\gamma\). This helps because the inferred probabilities \(P(y_{ik}=1|\Psi)\) might not necessarily reflect the true underlying phylogenetic signal, for example because the phylogenetic tree does not match the observed data well enough.

@@

\begin{equation}
  \label{eq:full}
  \begin{aligned}
   y_{ik} &\sim \text{Bernoulli}(\theta_{ik}) &\\
   \theta_{ik} &=
        \text{logit}^{-1}(\beta_{0_{k}} + \gamma_{k}\text{logit}(P(y_{ik}=1|\Psi)) & \\
        \hspace{-6cm} &~~~~ +\sum_{k=1}^{D}X_{ij}\beta_{\text{HLA}_{jk}}) & \\
    \beta_{0_{k}} &\sim \text{Normal}(0, 100^2) & \text{(*)}\\
   \gamma_{k} &\sim \text{Normal}(\mu_{\text{phy}}, \sigma_{\text{phy}}^2) & \\
    \mu_{\text{phy}} &\sim \text{Normal}(1, 1) & \text{(*)} \\
   \sigma_{\text{phy}} &\sim \text{Normal}^{+}(0, 0.5) & \text{(*)} \\
   \beta_{\text{epi}} &\sim \text{Normal}^{+}(1, 2) & \text{(*)} \\
   \beta_{\text{HLA}_{jk}} &\sim \text{Normal}(0, \tau_{k}^{2}\tilde{\lambda}^{2}_{jk}) & \\
   \tilde{\lambda}^{2}_{jk} &= \frac{c_{k}^{2}\lambda_{jk}^{2}}{c_{k}^{2} + \tau_{k}^{2}\lambda_{jk}^{2}} & \\
    c_{k}^2 &\sim \text{Inv-Gamma}(3.5, 3.5) & \text{(*)} \\
   \lambda_{jk} &\sim \text{Cauchy}^{+}(0, \sigma_{j}^2\text{exp}(Z_{jk}\beta_{\text{epi}})) & \\
   \tau_{k} &\sim \text{Cauchy}^{+}(0, \tau_{0k}) & \text{(*)} \\
   \tau_{0k} &= \frac{10}{D-10}\frac{2}{\sqrt{N}} & \\
  \end{aligned}
\end{equation}

@@

\begin{itemize}
  \item A large HIV dataset consisting of 1383 sequences, that were kindly provided by Zabrina Brumme and collaborators and were also used in the Phylogenetic Dependency Network study~\citep{Carlson2008}. The dataset contains sequences spanning the \textit{gag}, \textit{pol}, \textit{env}, \textit{nef}, \textit{vif}, \textit{vpr}, \textit{vpu}, \textit{tat} and \textit{rev} genes.
  \item A set of 351 HIV sequences mostly spanning the \textit{pol} gene from the Arevir database~\citep{Roomp2006}.
  \item A set of 544 Hepatitis-B-Virus sequences that were kindly provided by Jörg Timm. The dataset contains sequences of the HBe, LHBs, Pol and HBx proteins.
  \item A set of 104 Hepatitis-D-Virus sequences containing the HDV-antigen that were kindly provided by Michael Roggendorf. \hl{DaHo: bitte von Hadi geeignetes Zitat erfragen}
  \item A set of 41 HIV sequences spannign the \textit{gag} and \textit{pol} genes that were kindly provided by Rongge Yang. 
\end{itemize}
